---
title: "High-Risk HPV and Anal Cancer amongst IBD Patients"
format:
  html:
    toc: false
    toc-depth: 3
    toc-location: right
    echo: false
    css: styles/tables.css
    embed-resources: true
    self-contained-math: true
bibliography: references.bib
---

# Results Data

## HPV risk and odds math for IBD groups

This document lays out the proportion, risk ratio, odds, odds ratio, and confidence-interval math in Quarto-friendly Markdown. The outcome is a specific HPV diagnosis, and the exposure groups are non-IBD, Crohn's disease, and ulcerative colitis.[1][2]


## Raw counts

- Crohn's: 197 / 1,206,873
- UC: 319 / 1,191,972
- Non-IBD: 22,867 / 305,065,760


Crohn's
197/1,206,873  (0.016%)
(95% CI: +/- 0.002%, Z-Score: 1.96 Standard Error: 0.001)

UC
319/1,191,972 (0.027%)
95% Confidence Interval: +/- 0.003%
Z-Score: 1.96 Standard Error: 0.001

Non-IBD
22,867/305,065,760 (0.0075%)
95% Confidence Interval: +/- 0%
Z-Score: 1.96 Standard Error: 0

## Proportions

The sample proportion is:

$$
\hat{p} = \frac{x}{n}
$$

### Crohn's

$$
\hat{p}_{\text{Crohn}} = \frac{197}{1{,}206{,}873} \approx 0.0001633
$$

As a percent:

$$
0.0001633 \times 100 = 0.01633\%
$$

### UC

$$
\hat{p}_{\text{UC}} = \frac{319}{1{,}191{,}972} \approx 0.0002677
$$

As a percent:

$$
0.0002677 \times 100 = 0.02677\%
$$

### Non-IBD

$$
\hat{p}_{\text{non-IBD}} = \frac{22{,}867}{305{,}065{,}760} \approx 0.0000750
$$

As a percent:

$$
0.0000750 \times 100 = 0.00750\%
$$

Rounded to four decimal places in proportion form:

- Crohn's: 0.0002
- UC: 0.0003
- Non-IBD: 0.0001

These are all very small probabilities, so the rare disease assumption applies and the odds ratio closely approximates the risk ratio.[2][3]

## Risk ratios

Using non-IBD as the reference:

$$
RR_{\text{Crohn vs non-IBD}} = \frac{0.0001633}{0.0000750} \approx 2.1773
$$

$$
RR_{\text{UC vs non-IBD}} = \frac{0.0002677}{0.0000750} \approx 3.5693
$$

Interpretation:

- Crohn's has about 2.18 times the risk of HPV diagnosis compared with non-IBD.
- UC has about 3.57 times the risk of HPV diagnosis compared with non-IBD.

## Odds

Odds are defined as:

$$
\text{odds} = \frac{p}{1-p}
$$

### Non-IBD odds

$$
\text{odds}_{\text{non-IBD}} = \frac{0.0000750}{1 - 0.0000750} = \frac{0.0000750}{0.9999250} \approx 0.0000751
$$

### Crohn's odds

$$
\text{odds}_{\text{Crohn}} = \frac{0.0001633}{1 - 0.0001633} = \frac{0.0001633}{0.9998367} \approx 0.0001633
$$

### UC odds

$$
\text{odds}_{\text{UC}} = \frac{0.0002677}{1 - 0.0002677} = \frac{0.0002677}{0.9997323} \approx 0.0002678
$$

Because the probabilities are so small, each odds value is nearly identical to its corresponding risk.[1][2]

## Odds ratios

Using non-IBD as the reference:

$$
OR_{\text{Crohn vs non-IBD}} = \frac{0.0001633 / (1-0.0001633)}{0.0000750 / (1-0.0000750)} \approx \frac{0.0001633}{0.0000751} \approx 2.1770
$$

$$
OR_{\text{UC vs non-IBD}} = \frac{0.0002677 / (1-0.0002677)}{0.0000750 / (1-0.0000750)} \approx \frac{0.0002678}{0.0000751} \approx 3.5667
$$

Interpretation:

- Crohn's has about 2.18 times the odds of HPV diagnosis compared with non-IBD.
- UC has about 3.57 times the odds of HPV diagnosis compared with non-IBD.

Given how rare the outcome is, these ORs are essentially the same as the RRs in practical interpretation.[1][2][3]

## Standard error and 95% CI for a proportion

For a sample proportion, the standard error is:

$$
SE_{\hat{p}} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
$$

The Wald 95% confidence interval is:

$$
\hat{p} \pm 1.96 \times SE_{\hat{p}}
$$

This is the usual large-sample proportion CI formula.[4][5][6]

### Crohn's example

$$
SE_{\hat{p},\text{Crohn}} = \sqrt{\frac{0.0001633(1-0.0001633)}{1{,}206{,}873}} \approx 0.0000116
$$

$$
95\%\ CI = 0.0001633 \pm 1.96 \times 0.0000116
$$

$$
95\%\ CI \approx 0.0001633 \pm 0.0000227
$$

$$
95\%\ CI \approx (0.0001406,\ 0.0001860)
$$

As percentages:

$$
(0.01406\%,\ 0.01860\%)
$$

### UC example

$$
SE_{\hat{p},\text{UC}} = \sqrt{\frac{0.0002677(1-0.0002677)}{1{,}191{,}972}} \approx 0.0000150
$$

$$
95\%\ CI = 0.0002677 \pm 1.96 \times 0.0000150
$$

$$
95\%\ CI \approx 0.0002677 \pm 0.0000294
$$

$$
95\%\ CI \approx (0.0002383,\ 0.0002971)
$$

As percentages:

$$
(0.02383\%,\ 0.02971\%)
$$

### Non-IBD example

$$
SE_{\hat{p},\text{non-IBD}} = \sqrt{\frac{0.0000750(1-0.0000750)}{305{,}065{,}760}} \approx 0.0000005
$$

$$
95\%\ CI = 0.0000750 \pm 1.96 \times 0.0000005
$$

$$
95\%\ CI \approx 0.0000750 \pm 0.0000010
$$

$$
95\%\ CI \approx (0.0000741,\ 0.0000760)
$$

As percentages:

$$
(0.00741\%,\ 0.00760\%)
$$

## Manuscript-style wording

Possible wording:

> The incidence of HPV diagnosis was 0.00750% in the non-IBD group, 0.01633% in Crohn's disease, and 0.02677% in ulcerative colitis. Relative to non-IBD, this corresponds to an approximately 2.18-fold higher risk in Crohn's disease and a 3.57-fold higher risk in ulcerative colitis. Because the outcome was extremely rare, the corresponding odds ratios were numerically almost identical to the risk ratios.[1][2][3]

For logistic regression wording:

> Using non-IBD as the reference category, Crohn's disease and ulcerative colitis were associated with approximately 2.18-fold and 3.57-fold higher odds of HPV diagnosis, respectively. Given the very low absolute event rate, these odds ratios can be interpreted as closely approximating the corresponding relative risks.[1][2][3]

## Chi-square analysis

The chi-square analysis should use the raw counts, not the confidence intervals, Z scores, or standard errors. The confidence intervals describe uncertainty around each group's observed proportion. The chi-square test asks a different question: whether the distribution of HR-HPV positive vs HR-HPV negative patients differs by group.

### Step 1: load the reusable functions

The functions live in `scripts/csv-analysis.r`. When this Quarto file is rendered from the `qmd/` folder, the script is one directory up.

```{r}
#| label: chi-square-load-functions
#| echo: true
source("../scripts/csv-analysis.r")
```

### Step 2: read the CSV

This reads `data/hrhpv_group_counts.csv` and checks that the columns needed for the analysis are present.

```{r}
#| label: chi-square-read-counts
#| echo: true
counts <- read_hrhpv_counts()

counts |>
  dplyr::select(
    group,
    outcome_positive,
    outcome_negative,
    total,
    percent
  )
```

### Step 3: make the 3 by 2 contingency table

This is the raw table used by the overall Pearson chi-square test.

```{r}
#| label: chi-square-contingency-table
#| echo: true
contingency_table <- make_hrhpv_contingency_table(counts)

contingency_table
```

The rows are the three exposure groups. The columns are the two outcome states: HR-HPV positive and HR-HPV negative.

### Step 4: run the overall Pearson chi-square test

This test asks whether HR-HPV positivity is independent of group membership across all three groups.

```{r}
#| label: chi-square-overall-test
#| echo: true
overall_chi_square <- run_hrhpv_chi_square(
  counts = counts,
  correct = FALSE
)

overall_chi_square$summary_table
```

The `continuity_correction = FALSE` column means that Yates' continuity correction was not applied. That correction is mainly used for small 2 by 2 tables. Here, the counts are very large, so the standard Pearson chi-square test is appropriate.

The p value is extremely small. This is evidence against independence, not evidence for independence. In plain language, HR-HPV positivity differs by group.

Manuscript-style wording:

> HR-HPV positivity differed significantly by group on Pearson chi-square testing, X^2^(2) = 701, p < 0.001.

### Step 5: run the pairwise 2 by 2 chi-square tests

The overall chi-square test tells us that at least one group differs from another. Pairwise tests show which group comparisons are driving that result.

Here the three pairwise comparisons are:

- Crohn's disease vs non-IBD
- Ulcerative colitis vs non-IBD
- Crohn's disease vs ulcerative colitis

Because these are multiple follow-up tests, the p values should be adjusted. The code below uses Holm adjustment, which is a common conservative option.

```{r}
#| label: chi-square-pairwise-tests
#| echo: true
pairwise_chi_square <- run_hrhpv_pairwise_chi_square(
  counts = counts,
  comparisons = list(
    c("Crohn disease", "Non-IBD"),
    c("Ulcerative colitis", "Non-IBD"),
    c("Crohn disease", "Ulcerative colitis")
  ),
  correct = FALSE,
  p_adjust_method = "holm"
)

pairwise_chi_square
```

Interpretation:

- Crohn's disease differs significantly from non-IBD.
- Ulcerative colitis differs significantly from non-IBD.
- Crohn's disease also differs significantly from ulcerative colitis.

Because all adjusted p values are far below 0.05, the pairwise tests support the same overall conclusion: HR-HPV positivity is not independent of group status.
